# Importing externally imputed datasets

Note that we start this design doc with pseudo code parts, labelled clearly with `PSEUDOCODE`, which are not supposed to be running R code. This is done in order to first illustrate the ideas, without spending space on the implementation details. Only in the later end-to-end example we use executable R code.

## Objective

`rbmi` currently assumes that every analysis starts with `draws()` and
`impute()`. The objective of this change is to support completed datasets made
by another package or system while retaining the usual downstream workflow:

```text
PSEUDOCODE

imputations <- import(
    completed_datasets,
    vars = set_vars(
        subjid = "USUBJID",
        visit = "AVISIT",
        group = "TRT",
        outcome = "CHG",
        covariates = "BASE"
    )
)

analysis <- analyse(imputations, fun = ancova, vars = imputations$data$vars)
result <- pool(analysis)
```

The imported object should be a real `imputation` object. In particular,
`analyse()`, `extract_imputed_dfs()`, explicit delta adjustments, parallel
analysis and `pool()` should not need a separate external-data code path.

## Existing implementation in `rbmiUtils`

`rbmiUtils::analyse_mi_data()` takes one stacked data frame with an `IMPID`
column, splits it, calls the analysis function, and constructs an `analysis`
object itself. This demonstrates the use case, but duplicates validation,
delta handling, printing, and the construction logic from `rbmi::analyse()`.
It also bypasses the `imputation` interface, so improvements to that interface do not automatically benefit `analyse_mi_data()`.

The `rbmiUtils` development branch
`50-analyse_mi_data-when-the-exact-imputation-method-is-unknown` additionally
accepts a pooling rule and manufactures a Bayesian or conditional-mean method
object so that current `rbmi` validation passes. However, this creates a duplication of the interface between the `method` and the pooling rule. 

Strategically it makes sense to include `method` in the `import()` function signature, because it allows to completely reproduce the `imputed` object structure from the standard `rbmi` workflow. Typically in practice the programming team which produced the multiple imputed data sets will be the same or at least closely collaborating with the programmers who are writing the `rbmi` using code. Therefore it is known which statistical method was used with which assumptions, and can be passed transparently with the `method` argument. 

## Structure of an `rbmi` imputation object

An `rbmi` imputation currently contains:

- `imputations`: an `imputation_list_df` containing only the values that were
  missing in the source data;
- `data`: a `longdata` object containing the source data, variable names,
  missingness information and ICE metadata;
- `method`: the `method_*` object that determines the inference and pooling
  approach; and
- `references`: the reference-arm mapping used for reference-based imputation.

`analyse()` already goes through `extract_imputed_df()` to materialise each
completed data set. That is the appropriate starting point for our imported storage
backend. Converting completed data frames into the existing compact
`longdata` representation is not reliable without the original incomplete
data: completed data alone cannot reveal which cells were imputed, and two
imputations may legitimately contain the same imputed value. Therefore we do allow the 
optional passing of the `original` data set including the missing values. However,
this argument should not be required for the import to succeed.

## Proposed design

### New `import()` generic function

As with `draws()`, dispatch should be on the method, not on the data. The data
are always supplied in the same form: a non-empty list of completed data
frames. Method dispatch is useful because the expected number and layout of
those frames differs by inference method (in addition to the other strategic reasons mentioned above).

```text
PSEUDOCODE

import <- function(
    data,
    vars,
    method = NULL, # optional known method
    original = NULL, # optional original incomplete data set
    data_ice = NULL, # optional ICE metadata
    references = NULL, # optional reference-arm mapping
    source_id = NULL, # optional source ID for conditional-mean method
    ...
) {
    UseMethod("import", method)
}
```

The S3 methods are `import.bayes()`, `import.approxbayes()`,
`import.condmean()`, `import.bmlmi()` and `import.NULL()`.

For the last one, `UseMethod()` dispatches on the implicit class `"NULL"` when `method` is omitted. This means that the exact method that was used to generate the multiple imputation data sets in the first place is unknown, and only the standard Rubin's pooling method can be used downstream. 

These S3 methods perform
method-specific validation and then call one internal constructor, avoiding
duplication of storage code. This parallels `draws()`, whose methods establish
the sampling layout before returning a common `draws` object.

Here `data` is a non-empty list of completed data frames. The optional `original`
is the incomplete pre-imputation data set, and `data_ice` has the same meaning
as in `draws()`. Neither is needed for `analyse()` or `pool()`, but together
they enable truthful original-missingness and ICE metadata for
`delta_template()` downstream. `references` is optional provenance; it must not silently
default to an identity mapping, because that would claim knowledge about how
the external values were generated. 

`source_id` is only used in `import.condmean()`, where it names the
column that maps resampled subject occurrences back to subjects in `original`.

The implementation should require:

- all entries to be data frames containing the variables named by `vars`;
- a numeric, non-missing outcome in every completed data set;
- unique subject-occurrence/visit keys within every data set;
- invariant group values within each source subject and compatible column
  classes/factor levels (apart from occurrence IDs); and
- a data-set count compatible with `method`.

Rows can be sorted internally by subject and visit, so their input order need
not match where the method requires a common grid. Other columns are retained in full and may differ between
imputations; this permits externally imputed covariates even though `rbmi`'s own
imputation model only imputes the outcome. Thus also the completed data frames may contain additional columns beyond those required by `rbmi`.

### Method-specific layouts

The following table summarizes the number and layout of completed data sets for each import method, along with the corresponding pooling method:

| Import method | Number and layout of completed data sets | Pooling |
|---|---|---|
| `import.bayes()` | `method$n_samples`; common subject/visit grid | Rubin |
| `import.approxbayes()` | `method$n_samples`; common subject/visit grid, despite bootstrap being used to estimate model parameters | Rubin |
| `import.condmean(type = "bootstrap")` | original conditional-mean data set first, followed by `method$n_samples` bootstrap data sets | Bootstrap |
| `import.condmean(type = "jackknife")` | original conditional-mean data set first, followed by one leave-one-subject-out data set per original subject | Jackknife |
| `import.bmlmi()` | `method$B * method$D` data sets, ordered as `D` adjacent imputations within each bootstrap sample | BMLMI |
| `import.NULL()` | any positive number of data sets; common subject/visit grid | Rubin |

#### Bayes and Approximate Bayes

Bayes uses a fully Bayesian model to generate multiple imputations of the common subject/visit grid.
We therefore use a common subject/visit grid for all imputations.

Approximate Bayes uses bootstrap samples to estimate parameters, but each
completed data set is an imputation of the original subject grid. It therefore
uses the same fixed-grid validator as Bayes and Rubin's pooling method.

#### Conditional-mean bootstrap

Conditional-mean bootstrap is different: The first data set in the list represents the
original complete sample; every following data set in the list represents a bootstrap sample and
can contain the same source subject more than once. Within each completed data
set, `data[[vars$subjid]]` must identify *occurrences* uniquely. For example,
two selections of subject `101` could have occurrence IDs `b03_017` and
`b03_029`. The column named by `source_id` (only used in the `import.condmean()` method) contains `101` for both occurrences.
This explicit mapping avoids trying to infer bootstrap identity from duplicated
rows, and it supplies the `idmap` needed by existing delta handling.

For `method_condmean(type = "bootstrap")`, validation should require:

- `length(data) == method$n_samples + 1`;
- the original sample data set to have one occurrence for every original subject;
- every bootstrap data set to have the same number of subject occurrences as
  the original sample, each with a complete visit grid;
- all `source_id` values to occur in `original`, or in the first completed
  data set when `original` is unavailable; and
- when strata are available, each bootstrap sample to retain the original
  number of occurrences per stratum, matching `rbmi`'s stratified bootstrap.

#### Jackknife

For jackknife, validation should require `N + 1` data sets for `N` original
subjects, with the original data set first and each later data set omitting
exactly one distinct source subject. For BMLMI, `source_id` has the same meaning
as for conditional-mean bootstrap; each adjacent block of `D` data sets must
have an identical subject-occurrence map. The full data frames within a block
may differ in their imputed outcomes. So that means
`import.bmlmi()` follows the same pattern as `import.condmean()`, using a
helper that checks `B * D` data sets and identical
occurrence maps within each block of `D`. 

We note that there is deliberately no
`import.data.frame()` method: dispatch belongs to `method`, while a separate
convenience helper function could split stacked `IMPID` data before calling `import()`.

### Import without a specified method

When `method = NULL` is left as in the default, the import assumes ordinary Rubin-style multiple imputation.
There is no user-facing method constructor or pooling choice: a different
pooling rule would need information about the imputation process that the completed
data sets alone cannot supply. Internally, `import.NULL()` validates the fixed
grid and records the inferred number of imputations in a private method object,
so the existing `imputation` and `analysis` interfaces retain their method
metadata. This object is created only by `import()` and is not accepted by
`draws()`. It can also not be confused by the user because it is not a `method_*()` created object.

```text
PSEUDOCODE

import.NULL <- function(data, vars, method, ...) {
    validate_fixed_grid(data, vars, expected = length(data))
    # Create private method object here:
    imported_method <- structure(
        list(n_samples = length(data)),
        class = c("method", "unknown")
    )
    import_completed(data, vars, imported_method, ...)
}
```

`validate.analysis()` can then use the recorded `n_samples` as it does for
other methods. The internal `unknown` class denotes provenance and pooling
behaviour, not a new imputation algorithm.

### Internal pooling look-up

The package now includes the internal, roxygen2-documented
`pooling_method()` helper in `R/analyse.R`, with tests for every method mapping.
Both `as_analysis()` and `validate.analysis()` use it in place of inspecting
`class(method)[[2]]`. This is the package change needed for the private
`unknown` method in the executable prototype. Its design is:

```text
PSEUDOCODE

pooling_method <- function(method) {
    if (inherits(method, "unknown")) {
        return("rubin")
    }
    if (inherits(method, "bayes") || inherits(method, "approxbayes")) {
        return("rubin")
    }
    if (inherits(method, "condmean")) {
        return(if (method$type == "jackknife") "jackknife" else "bootstrap")
    }
    if (inherits(method, "bmlmi")) {
        return("bmlmi")
    }
    stop("Unsupported method object")
}
```

This removes the assumption that the decisive method class is always the
second class. `mcse()` should continue to accept only actual
`bayes` and `approxbayes` methods. No special `analyse()` method or wrapper is
needed: `analyse()` remains method-agnostic, and `as_analysis()` uses
`pooling_method()` when it assigns the class of the collected results.

### Storage behind the imputation interface

Keep the top-level `imputation` shape, but allow two validated storage pairs:

| Source | `imputations` | `data` metadata/backend |
|---|---|---|
| `impute()` | compact `imputation_list_df` | existing `longdata` |
| `import()` | `completed_imputation_list` | new `completed_data` |

Each `completed_imputation` owns one complete data frame and its
occurrence-to-source ID map. `completed_data` stores `vars` plus optional
metadata derived from `original`. A small internal generic materialises either
representation:

```text
PSEUDOCODE

materialise_imputation <- function(x, data, idmap = FALSE) {
    UseMethod("materialise_imputation")
}

materialise_imputation.imputation_df <- function(x, data, idmap = FALSE) {
    data$get_data(x, idmap = idmap)
}

materialise_imputation.completed_imputation <- function(
    x, data, idmap = FALSE
) {
    out <- x$data
    # Apply the same deterministic subject-ID remapping used by longdata.
    # x$source_ids maps each occurrence to an original subject.
    remap_subject_ids(
        out,
        vars = data$vars,
        source_ids = x$source_ids,
        idmap = idmap
    )
}
```

`extract_imputed_df()` then calls this generic and retains its existing common
delta-adjustment logic. Consequently `analyse()` and
`extract_imputed_dfs()` do not care where the completed data originated.
`validate.imputation()` validates the two supported storage pairs and rejects
mixed representations.

This is preferable to adding `if (imported)` logic to `analyse()`: storage is
resolved at the existing materialisation boundary, and all analyses still
consume ordinary data frames through the same path.

### Metadata-dependent behaviour

If `original` is supplied, `completed_data` records the original missingness
mask. On a fixed grid it should verify identical subject/visit keys and
invariant observed outcomes across all completed data sets. 

If `data_ice` is also supplied, it records strategy and post-ICE metadata using the existing validation rules.
Without these inputs, the corresponding metadata are unknowable and must be
represented as unavailable, not guessed.

For resampled data, the invariance check is performed after joining each
occurrence to `original` through `source_id`. The occurrence IDs are used only
inside the completed data set; delta templates and provenance remain keyed by
the source subject. This mirrors the ID scrambling and `idmap` behaviour of
the existing compact backend.

- `analyse(..., delta = explicit_data_frame)` remains available, because the
  caller explicitly identifies the rows and offsets.
- `delta_template()` should give an informative error unless `original` was
  supplied. As in `draws()`, a missing `data_ice` means that no ICE metadata is
  applied; supplying it enables the post-ICE and strategy-derived fields.
- `print.imputation()` should show `Original missingness: unavailable` and
  `References: not supplied` where appropriate.
- `extract_imputed_dfs()` must reproduce the imported data (apart from rbmi's
  documented subject-ID remapping and row normalisation).

## Testing and implementation sequence

This is a rough roadmap how the functionality according to the above design 
can be implemented efficiently in the package:

1. Done (was necessary to get below prototype code to run): add `pooling_method()`, including
   the private unknown method; use it in analysis construction and validation.
2. Add complete test coverage for `pooling_method()`.
3. Add the completed-data classes, validators and materialisation methods.
4. Implement the method-dispatched `import()` methods and round-trip tests
   comparing each imported data set with `extract_imputed_dfs()`.
5. Test conditional-mean bootstrap and jackknife layouts, including repeated
   source subjects, occurrence ID remapping, ordering and method counts. Test
   BMLMI blocks separately.
6. Test failures for inconsistent keys, duplicate occurrence/visit rows,
   factor level drift, remaining outcome missingness, invalid source IDs and
   incompatible method counts.
7. Update printing and document the unavailable-metadata behaviour.
8. Test `import() -> analyse() -> pool()` against direct per-data-set analyses,
   including custom analysis functions, explicit delta data and parallel use.
9. Test conditional-mean bootstrap and jackknife layouts, including repeated
   source subjects, occurrence ID remapping, ordering and method counts. Test
   BMLMI blocks separately.
10. Test failures for inconsistent keys, duplicate occurrence/visit rows,
   factor level drift, remaining outcome missingness, invalid source IDs and
   incompatible method counts.
11. Update printing and document the unavailable-metadata behaviour.


## End-to-end examples


### Prototype code

The
new `pooling_method()` helper in the package lets `rbmi::analyse()` recognize the
private unknown method and select Rubin pooling.

The prototype uses `rbmi`'s existing compact imputation storage. In the actual implementation this may be replaced with the above proposed storage scheme.
When `original` is omitted, this compact-storage prototype builds a source
grid from the union of all completed data sets, matching bootstrap occurrences
by `source_id`. Non-outcome fields must agree for each subject and visit. It
then marks every outcome as missing so each completed outcome can be stored.
Its missingness metadata therefore does not describe the original data; the user can supply
`original` when that metadata matters.
Omitting `references` records an empty mapping, not an inferred reference arm.

```{r}
#| label: executable-import-prototype
#| warning: false

library(rbmi)

# Dispatch imports by the method used to create the completed data sets.
import <- function(
    data,
    vars,
    method = NULL,
    original = NULL,
    data_ice = NULL,
    references = NULL,
    source_id = NULL,
    ...
) {
    UseMethod("import", method)
}

# Check the fields and outcome values required by every import method.
validate_completed <- function(data, vars) {
    stopifnot(
        length(data) > 0L,
        all(vapply(data, is.data.frame, logical(1))),
        all(vapply(
            data,
            function(x) {
                all(c(vars$subjid, vars$visit, vars$group, vars$outcome) %in%
                    names(x)) &&
                    is.numeric(x[[vars$outcome]]) &&
                    !anyNA(x[[vars$outcome]]) &&
                    !anyDuplicated(x[c(vars$subjid, vars$visit)])
            },
            logical(1)
        ))
    )
}

# Require the expected number of data sets and the same subject-visit grid.
validate_fixed_grid <- function(data, vars, expected) {
    validate_completed(data, vars)
    stopifnot(length(data) == expected)
    # Compare keys independently of row order.
    key <- function(x) {
        sort(paste(x[[vars$subjid]], x[[vars$visit]], sep = "\r"))
    }
    keys <- lapply(data, key)
    stopifnot(all(vapply(keys[-1L], identical, logical(1), keys[[1L]])))
}

# Reconstruct source rows from completed data when the original is unavailable.
# Map bootstrap occurrences to their source subjects before deduplicating.
source_from_completed <- function(data, vars, source_id) {
    source_rows <- do.call(rbind, lapply(data, function(x) {
        if (!is.null(source_id)) x[[vars$subjid]] <- x[[source_id]]
        x[setdiff(names(x), c(vars$outcome, source_id))]
    }))
    original <- unique(source_rows)
    stopifnot(!anyDuplicated(original[c(vars$subjid, vars$visit)]))
    original[[vars$subjid]] <- factor(original[[vars$subjid]])
    original[[vars$outcome]] <- 0
    original
}

# Store one completed data set as values for the source's missing visits.
# Preserve the source ID for each sampled bootstrap occurrence.
to_imputation_df <- function(x, vars, longdata, source_id) {
    occurrences <- unique(as.character(x[[vars$subjid]]))
    singles <- lapply(occurrences, function(occurrence) {
        rows <- x[as.character(x[[vars$subjid]]) == occurrence, , drop = FALSE]
        source <- if (is.null(source_id)) {
            occurrence
        } else {
            unique(as.character(rows[[source_id]]))
        }
        stopifnot(length(source) == 1L, source %in% longdata$ids)

        visit_order <- match(
            longdata$visits,
            as.character(rows[[vars$visit]])
        )
        stopifnot(!anyNA(visit_order))
        values <- rows[[vars$outcome]][visit_order]
        rbmi:::imputation_single(
            id = source,
            values = values[longdata$is_missing[[source]]]
        )
    })
    rbmi:::imputation_df(singles)
}

# Build the compact rbmi object shared by the method-specific import methods.
import_compact <- function(
    data,
    vars,
    method,
    original,
    data_ice,
    references,
    source_id
) {
    no_original <- is.null(original)
    if (no_original) {
        original <- source_from_completed(data, vars, source_id)
    }
    longdata <- longDataConstructor$new(original, vars)
    longdata$set_strategies(data_ice)
    if (no_original) {
        longdata$data[[vars$outcome]] <- NA_real_
        longdata$values <- lapply(longdata$values, function(x) rep(NA_real_, length(x)))
        longdata$is_missing <- lapply(longdata$is_missing, function(x) rep(TRUE, length(x)))
    }

    stored <- lapply(data, to_imputation_df,
        vars = vars, longdata = longdata, source_id = source_id
    )
    stored <- do.call(rbmi:::imputation_list_df, stored)
    if (is.null(references)) {
        references <- setNames(character(), character())
    }
    references <- add_class(references, "references")

    out <- rbmi:::as_imputation(
        imputations = stored,
        data = longdata,
        method = method,
        references = references
    )
    validate(out)
    out
}

# Check Bayesian sample layout before using the shared compact constructor.
import.bayes <- function(
    data, vars, method, original = NULL, data_ice = NULL, references = NULL,
    source_id = NULL, ...
) {
    validate_fixed_grid(data, vars, method$n_samples)
    import_compact(
        data, vars, method, original, data_ice, references, source_id
    )
}

# Approximate Bayesian imports have the same layout as Bayesian imports.
import.approxbayes <- import.bayes

# Record unknown provenance while retaining the number of imputations.
import.NULL <- function(
    data, vars, method, original = NULL, data_ice = NULL, references = NULL,
    source_id = NULL, ...
) {
    validate_fixed_grid(data, vars, length(data))
    stopifnot(is.null(source_id))
    imported_method <- structure(
        list(n_samples = length(data)),
        class = c("method", "unknown")
    )
    import_compact(
        data, vars, imported_method, original, data_ice, references, source_id
    )
}

# Check bootstrap or jackknife layout and map occurrences to source IDs.
import.condmean <- function(
    data, vars, method, original = NULL, data_ice = NULL, references = NULL,
    source_id = NULL, ...
) {
    validate_completed(data, vars)
    stopifnot(
        method$type %in% c("bootstrap", "jackknife"),
        !is.null(source_id),
        all(vapply(data, function(x) source_id %in% names(x), logical(1)))
    )
    source_sets <- lapply(data, function(x) {
        by_occurrence <- split(
            as.character(x[[source_id]]), as.character(x[[vars$subjid]])
        )
        stopifnot(all(lengths(lapply(by_occurrence, unique)) == 1L))
        vapply(by_occurrence, function(ids) ids[[1L]], character(1))
    })
    originals <- unname(source_sets[[1L]])
    stopifnot(!anyDuplicated(originals))
    if (method$type == "bootstrap") {
        stopifnot(
            length(data) == method$n_samples + 1L,
            all(lengths(source_sets[-1L]) == length(originals)),
            all(vapply(source_sets[-1L], function(ids) {
                all(ids %in% originals)
            }, logical(1)))
        )
    } else {
        stopifnot(length(data) == length(originals) + 1L)
        omitted <- vapply(source_sets[-1L], function(ids) {
            missing <- setdiff(originals, ids)
            stopifnot(length(missing) == 1L, setequal(ids, setdiff(originals, missing)))
            missing
        }, character(1))
        stopifnot(setequal(omitted, originals))
    }
    import_compact(
        data, vars, method, original, data_ice, references, source_id
    )
}

# Keep the D adjacent imputations for each BMLMI bootstrap occurrence map together.
import.bmlmi <- function(
    data, vars, method, original = NULL, data_ice = NULL, references = NULL,
    source_id = NULL, ...
) {
    validate_completed(data, vars)
    stopifnot(
        length(data) == method$B * method$D,
        !is.null(source_id),
        all(vapply(data, function(x) source_id %in% names(x), logical(1)))
    )
    occurrence_map <- function(x) {
        ids <- split(as.character(x[[source_id]]), as.character(x[[vars$subjid]]))
        stopifnot(all(lengths(lapply(ids, unique)) == 1L))
        vapply(ids, function(y) y[[1L]], character(1))
    }
    maps <- lapply(data, occurrence_map)
    for (b in seq_len(method$B)) {
        block <- ((b - 1L) * method$D + 1L):(b * method$D)
        stopifnot(all(vapply(maps[block], identical, logical(1), maps[[block[[1L]]]])))
    }
    import_compact(
        data, vars, method, original, data_ice, references, source_id
    )
}
```

The first executed example imports ordinary MI data with unknown imputation
provenance and uses Rubin pooling. The second example shows the additional ID
information required for conditional-mean bootstrap data.

### Rubin's pooling for unknown `method`

This is the complete user workflow when it is not clear which multiple imputation method was used. Only Rubin's pooling can be used in this case.

```{r}
#| label: example-rubin

# First let's create the original dataset with missing values.
set.seed(2026)
n_subjects <- 40L
n_imputations <- 5L

subject <- data.frame(
    USUBJID = factor(sprintf("P%03d", seq_len(n_subjects))),
    ARM = factor(rep(c("Placebo", "Active"), each = n_subjects / 2L)),
    BASE = rnorm(n_subjects, mean = 10, sd = 2)
)

original <- subject[rep(seq_len(n_subjects), each = 2L), ]
original$AVISIT <- factor(
    rep(c("Week 4", "Week 8"), times = n_subjects),
    levels = c("Week 4", "Week 8")
)
original$CHG <-
    -0.4 * original$BASE +
    1.5 * (original$ARM == "Active") +
    0.5 * (original$AVISIT == "Week 8") +
    rnorm(nrow(original))

missing_subjects <- sample(levels(original$USUBJID), 10L)
is_missing <-
    original$USUBJID %in% missing_subjects &
    original$AVISIT == "Week 8"
original$CHG[is_missing] <- NA_real_

# Now we create the data sets representing the multiple imputations.
completed <- lapply(seq_len(n_imputations), function(i) {
    out <- original
    out$CHG[is_missing] <-
        -0.4 * out$BASE[is_missing] +
        1.5 * (out$ARM[is_missing] == "Active") +
        0.5 +
        rnorm(sum(is_missing))
    out
})

# Finally here comes the actual user interface demo.
vars <- set_vars(
    subjid = "USUBJID",
    visit = "AVISIT",
    group = "ARM",
    outcome = "CHG",
    covariates = "BASE"
)

imputations <- import(
    data = completed,
    vars = vars
)

analysis <- rbmi::analyse(
    imputations = imputations,
    fun = ancova,
    vars = vars,
    visits = "Week 8"
)
pooled <- pool(analysis)
as.data.frame(pooled)
```

We can see in the above example that the import is easy and brings us back to the standard `rbmi` workflow afterwards.


### Conditional-mean bootstrap

Let's look also at a more complicated example with conditional mean imputation.
The original-sample conditional means are first in `completed_bootstrap`.
Every later element is one stratified bootstrap sample. `USUBJID` identifies a
unique sampled occurrence, while `SOURCE_ID` retains the original subject ID;
therefore a source subject selected twice is represented unambiguously.

```{r}
#| label: example-condmean-bootstrap

B <- 10L

# For illustration, create deterministic conditional-mean replacements.
conditional_mean <- original
conditional_mean$CHG[is_missing] <-
    -0.4 * conditional_mean$BASE[is_missing] +
    1.5 * (conditional_mean$ARM[is_missing] == "Active") +
    0.5
conditional_mean$SOURCE_ID <- as.character(conditional_mean$USUBJID)

arm_by_subject <- split(
    as.character(subject$USUBJID),
    subject$ARM
)

bootstrap_data <- lapply(seq_len(B), function(b) {
    sampled_sources <- unlist(lapply(
        arm_by_subject,
        function(ids) sample(ids, length(ids), replace = TRUE)
    ), use.names = FALSE)

    sampled_occurrences <- lapply(seq_along(sampled_sources), function(j) {
        source <- sampled_sources[[j]]
        out <- conditional_mean[
            as.character(conditional_mean$USUBJID) == source,
            ,
            drop = FALSE
        ]
        out$SOURCE_ID <- source
        out$USUBJID <- sprintf("B%02d_O%03d", b, j)
        out
    })

    out <- do.call(rbind, sampled_occurrences)
    rownames(out) <- NULL
    out$USUBJID <- factor(out$USUBJID)
    out
})

completed_bootstrap <- c(list(conditional_mean), bootstrap_data)

# Here we need to use the method and source_id arguments.
bootstrap_imputations <- import(
    data = completed_bootstrap,
    vars = vars,
    method = method_condmean(type = "bootstrap", n_samples = B),
    source_id = "SOURCE_ID"
)

bootstrap_analysis <- rbmi::analyse(
    imputations = bootstrap_imputations,
    fun = ancova,
    vars = vars,
    visits = "Week 8"
)
bootstrap_pooled <- pool(bootstrap_analysis)
as.data.frame(bootstrap_pooled)
```

Here `pool()` defaults to `type = "percentile"` conditional-mean bootstrap pooling, therefore there are no standard errors reported. In this case all p-values are zero because all 10 bootstrap samples show a positive effect. Alternatively we could use normal approximation standard errors and p-values:

```{r}
bootstrap_pooled <- pool(bootstrap_analysis, type = "normal")
as.data.frame(bootstrap_pooled)
```
