# Importing externally imputed datasets

## Objective

`rbmi` currently assumes that every analysis starts with `draws()` and
`impute()`. The objective of this change is to support completed datasets made
by another package or system while retaining the usual downstream workflow:

```r
imputations <- import(
    completed_datasets,
    vars = set_vars(
        subjid = "USUBJID",
        visit = "AVISIT",
        group = "TRT",
        outcome = "CHG",
        covariates = "BASE"
    ),
    method = method_unknown(pooling = "rubin")
)

analysis <- analyse(imputations, fun = ancova, vars = imputations$data$vars)
result <- pool(analysis)
```

The imported object should be a real `imputation` object. In particular,
`analyse()`, `extract_imputed_dfs()`, explicit delta adjustments, parallel
analysis and `pool()` should not need a separate external-data code path.

## Existing implementation in `rbmiUtils`

`rbmiUtils::analyse_mi_data()` takes one stacked data frame with an `IMPID`
column, splits it, calls the analysis function, and constructs an `analysis`
object itself. This demonstrates the use case, but duplicates validation,
delta handling, printing, and the construction logic from `rbmi::analyse()`.
It also bypasses the `imputation` interface, so improvements to that interface
do not automatically benefit `analyse_mi_data()`.

The `rbmiUtils` development branch
`50-analyse_mi_data-when-the-exact-imputation-method-is-unknown` additionally
accepts a pooling rule and manufactures a Bayesian or conditional-mean method
object so that current `rbmi` validation passes. Its important insight is that
the pooling rule, rather than the exact imputation algorithm, is what the
analysis and pooling steps need. Manufacturing a known method is nevertheless
misleading: for example, an external Rubin-style MI data set was not thereby
created by `method_bayes()`.

## Structure of an `rbmi` imputation object

An `rbmi` imputation currently contains:

```text
imputation
+-- imputations: imputation_list_df (only the values missing in the source data)
+-- data:        longdata (source data, variable names, missingness and ICE metadata)
+-- method:      method_* object
`-- references:  reference-arm mapping
```

`analyse()` already goes through `extract_imputed_df()` to materialise each
completed data set. That is the appropriate starting point for our imported storage
backend. Converting completed data frames into the existing compact
`longdata` representation is not reliable without the original incomplete
data: completed data alone cannot reveal which cells were imputed, and two
imputations may legitimately contain the same imputed value. Treating every
outcome as originally missing would also make `print()` and
`delta_template()` incorrect.

## Proposed design

### New `import()` generic function

As with `draws()`, dispatch should be on the method, not on the data. The data
are always supplied in the same form: a non-empty list of completed data
frames. Method dispatch is useful because the expected number and layout of
those frames differs by inference method.

```r
import <- function(
    data,
    vars,
    method = method_unknown(),
    original = NULL,
    data_ice = NULL,
    references = NULL,
    source_id = NULL,
    ...
) {
    UseMethod("import", method)
}

import.unknown <- function(
    data,
    vars,
    method,
    original = NULL,
    data_ice = NULL,
    references = NULL,
    ...
) {
    # Validate and construct an imputation object.
}
```

The methods are `import.bayes()`, `import.approxbayes()`,
`import.condmean()`, `import.bmlmi()` and `import.unknown()`. They perform
method-specific validation and then call one internal constructor, avoiding
duplication of storage code. This parallels `draws()`, whose methods establish
the sampling layout before returning a common `draws` object.

Here `data` is a non-empty list of completed data frames. The optional `original`
is the incomplete pre-imputation data set, and `data_ice` has the same meaning
as in `draws()`. Neither is needed for `analyse()` or `pool()`, but together
they enable truthful original-missingness and ICE metadata for
`delta_template()` downstream. `references` is optional provenance; it must not silently
default to an identity mapping, because that would claim knowledge about how
the external values were generated. 

`source_id` is only used in `import.condmean()`, where it names the
column that maps resampled subject occurrences back to subjects in `original`.

Because other packages also export a
function named `import()`, documentation should recommend `rbmi::import()` in
ambiguous sessions.

The first implementation should require:

- all entries to be data frames containing the variables named by `vars`;
- a numeric, non-missing outcome in every completed data set;
- unique subject-occurrence/visit keys within every data set;
- invariant group values within each source subject and compatible column
  classes/factor levels (apart from occurrence IDs); and
- a data-set count compatible with `method`.

Rows can be sorted internally by subject and visit, so their input order need
not match where the method requires a common grid. Other columns are retained in full and may differ between
imputations; this permits externally imputed covariates even though `rbmi`'s own
imputation model only imputes the outcome. 

### Method-specific layouts

The following contracts are part of the import, not assumptions left to
`analyse()`:

| Import method | Number and layout of completed data sets | Pooling |
|---|---|---|
| `import.bayes()` | `method$n_samples`; common subject/visit grid | Rubin |
| `import.approxbayes()` | `method$n_samples`; common subject/visit grid, despite bootstrap being used to estimate model parameters | Rubin |
| `import.condmean(type = "bootstrap")` | original conditional-mean data set first, followed by `method$n_samples` bootstrap data sets | Bootstrap |
| `import.condmean(type = "jackknife")` | original conditional-mean data set first, followed by one leave-one-subject-out data set per original subject | Jackknife |
| `import.bmlmi()` | `method$B * method$D` data sets, ordered as `D` adjacent imputations within each bootstrap sample | BMLMI |
| `import.unknown()` | `method$n_samples`; common subject/visit grid | Rubin |

Approximate Bayes uses bootstrap samples to estimate parameters, but each
completed data set is an imputation of the original subject grid. It therefore
uses the same fixed-grid validator as Bayes and unknown Rubin-style MI.

Conditional-mean bootstrap is different. The first data set in the list represents the
original complete sample; every following data set in the list represents a bootstrap sample and
can contain the same source subject more than once. Within each completed data
set, `data[[vars$subjid]]` must identify *occurrences* uniquely. For example,
two selections of subject `101` could have occurrence IDs `b03_017` and
`b03_029`. The column named by `source_id` (only used in the `import.condmean()` method) contains `101` for both occurrences.
This explicit mapping avoids trying to infer bootstrap identity from duplicated
rows, and it supplies the `idmap` needed by existing delta handling.

For `method_condmean(type = "bootstrap")`, validation should require:

- `length(data) == method$n_samples + 1`;
- the original sample data set to have one occurrence for every original subject;
- every bootstrap data set to have the same number of subject occurrences as
  the original sample, each with a complete visit grid;
- all `source_id` values to occur in `original`, or in the first completed
  data set when `original` is unavailable; and
- when strata are available, each bootstrap sample to retain the original
  number of occurrences per stratum, matching `rbmi`'s stratified bootstrap.

For jackknife, validation should require `N + 1` data sets for `N` original
subjects, with the original data set first and each later data set omitting
exactly one distinct source subject. For BMLMI, `source_id` has the same meaning
as for conditional-mean bootstrap; each adjacent block of `D` data sets must
have an identical subject-occurrence map. The full data frames within a block
may differ in their imputed outcomes.

### Unknown method

We add a genuine method class instead of disguising the pooling choice as an
existing method. We can only offer Rubin's pooling method here, as we would need to 
have other information about the imputation process to justify using a different pooling method.

```r
method_unknown <- function() {
    structure(
        list(pooling = "Rubin", n_samples = NA_integer_),
        class = c("method", "unknown")
    )
}
```

`import.unknown()` fills `n_samples` after seeing the data, matching the convention currently
expected by `validate.analysis()`. 

Unknown Rubin-style MI uses the fixed-grid validator. 

### Internal pooling look-up

Introduce one internal helper, used by both `as_analysis()` and validation,
instead of inspecting `class(method)[[2]]` in several places:

```r
pooling_method <- function(method) {
    if (inherits(method, "unknown")) {
        return(method$pooling)
    }
    if (inherits(method, "bayes") || inherits(method, "approxbayes")) {
        return("rubin")
    }
    if (inherits(method, "condmean")) {
        return(if (method$type == "jackknife") "jackknife" else "bootstrap")
    }
    if (inherits(method, "bmlmi")) {
        return("bmlmi")
    }
    stop("Unsupported method object")
}
```

This also removes the fragile assumption that the decisive method class is
always the second class. `mcse()` should continue to accept only actual
`bayes` and `approxbayes` methods. No `analyse_unknown()` method or wrapper is
needed: `analyse()` remains method-agnostic, and `as_analysis()` uses
`pooling_method()` when it assigns the class of the collected results.

### Storage behind the imputation interface

Keep the top-level `imputation` shape, but allow two validated storage pairs:

| Source | `imputations` | `data` metadata/backend |
|---|---|---|
| `impute()` | compact `imputation_list_df` | existing `longdata` |
| `import()` | `completed_imputation_list` | new `completed_data` |

Each `completed_imputation` owns one complete data frame and its
occurrence-to-source ID map. `completed_data` stores `vars` plus optional
metadata derived from `original`. A small internal generic materialises either
representation:

```r
materialise_imputation <- function(x, data, idmap = FALSE) {
    UseMethod("materialise_imputation")
}

materialise_imputation.imputation_df <- function(x, data, idmap = FALSE) {
    data$get_data(x, idmap = idmap)
}

materialise_imputation.completed_imputation <- function(
    x, data, idmap = FALSE
) {
    out <- x$data
    # Apply the same deterministic subject-ID remapping used by longdata.
    # x$source_ids maps each occurrence to an original subject.
    remap_subject_ids(
        out,
        vars = data$vars,
        source_ids = x$source_ids,
        idmap = idmap
    )
}
```

`extract_imputed_df()` then calls this generic and retains its existing common
delta-adjustment logic. Consequently `analyse()` and
`extract_imputed_dfs()` do not care where the completed data originated.
`validate.imputation()` validates the two supported storage pairs and rejects
mixed representations.

This is preferable to adding `if (imported)` logic to `analyse()`: storage is
resolved at the existing materialisation boundary, and all analyses still
consume ordinary data frames through the same path.

### Metadata-dependent behaviour

If `original` is supplied, `completed_data` records the original missingness
mask. On a fixed grid it should verify identical subject/visit keys and
invariant observed outcomes across all completed data sets. If `data_ice` is also supplied, it
records strategy and post-ICE metadata using the existing validation rules.
Without these inputs, the corresponding metadata are unknowable and must be
represented as unavailable, not guessed.

For resampled data, the invariance check is performed after joining each
occurrence to `original` through `source_id`. The occurrence IDs are used only
inside the completed data set; delta templates and provenance remain keyed by
the source subject. This mirrors the ID scrambling and `idmap` behaviour of
the existing compact backend.

- `analyse(..., delta = explicit_data_frame)` remains available, because the
  caller explicitly identifies the rows and offsets.
- `delta_template()` should give an informative error unless `original` was
  supplied. As in `draws()`, a missing `data_ice` means that no ICE metadata is
  applied; supplying it enables the post-ICE and strategy-derived fields.
- `print.imputation()` should show `Original missingness: unavailable` and
  `References: not supplied` where appropriate.
- `extract_imputed_dfs()` must reproduce the imported data (apart from rbmi's
  documented subject-ID remapping and row normalisation).

## Prototype constructor

The following illustrates dispatch and the conditional-mean checks. The class
constructors and error helpers would be internal in the production
implementation.

```r
import <- function(
    data,
    vars,
    method = method_unknown(),
    original = NULL,
    data_ice = NULL,
    references = NULL,
    source_id = NULL,
    ...
) {
    UseMethod("import", method)
}

import.bayes <- function(data, vars, method, ...) {
    validate_fixed_grid(data, vars, expected = method$n_samples)
    import_completed(data, vars, method, ...)
}

import.approxbayes <- import.bayes

import.condmean <- function(
    data,
    vars,
    method,
    original = NULL,
    data_ice = NULL,
    references = NULL,
    source_id = NULL,
    ...
) {
    if (method$type == "bootstrap") {
        validate_bootstrap_layout(
            data = data,
            original = original,
            vars = vars,
            source_id = source_id,
            expected_replicates = method$n_samples
        )
    } else {
        validate_jackknife_layout(
            data = data,
            original = original,
            vars = vars,
            source_id = source_id
        )
    }

    import_completed(
        data = data,
        vars = vars,
        method = method,
        original = original,
        data_ice = data_ice,
        references = references,
        source_id = source_id
    )
}

import.unknown <- function(data, vars, method, ...) {
    method$n_samples <- switch(
        method$pooling,
        rubin = length(data),
        bootstrap = length(data) - 1L,
        NULL
    )
    validate_unknown_layout(data, vars, method, ...)
    import_completed(data, vars, method, ...)
}

import_completed <- function(
    data,
    vars,
    method,
    original = NULL,
    data_ice = NULL,
    references = NULL,
    source_id = NULL
) {
    validate(vars)
    stopifnot(
        length(data) > 0L,
        all(vapply(data, is.data.frame, logical(1)))
    )

    required <- unique(c(
        vars$subjid, vars$visit, vars$group, vars$outcome,
        vars$strata, extract_covariates(vars$covariates)
    ))
    stopifnot(all(vapply(
        data,
        function(x) all(required %in% names(x)),
        logical(1)
    )))

    stored <- lapply(data, function(x) {
        source <- if (is.null(source_id)) x[[vars$subjid]] else x[[source_id]]
        id_pairs <- unique(data.frame(
            occurrence = as.character(x[[vars$subjid]]),
            source = as.character(source)
        ))
        stopifnot(!anyDuplicated(id_pairs$occurrence))
        source_map <- setNames(id_pairs$source, id_pairs$occurrence)
        structure(
            list(data = x, source_ids = source_map),
            class = c("completed_imputation", "list")
        )
    })
    class(stored) <- c("completed_imputation_list", "list")

    metadata <- new_completed_data(
        vars = vars,
        original = original,
        data_ice = data_ice,
        prototype = data[[1L]]
    )

    out <- as_imputation(
        imputations = stored,
        data = metadata,
        method = method,
        references = references
    )
    validate(out)
    out
}
```

`import.bmlmi()` follows the same pattern as `import.condmean()`, using a
`validate_bmlmi_layout()` helper that checks `B * D` data sets and identical
occurrence maps within each block of `D`. There is deliberately no

`import.data.frame()` method: dispatch belongs to `method`, while a separate
convenience helper function could split stacked `IMPID` data before calling `import()`.

## Testing and implementation sequence

1. Add `method_unknown()`, `pooling_method()` and tests for every pooling
   mapping; refactor existing methods without changing their results.
2. Add the completed-data classes, validators and materialisation methods.
3. Implement the method-dispatched `import()` methods and round-trip tests
   comparing each imported data set with `extract_imputed_dfs()`.
4. Test `import() -> analyse() -> pool()` against direct per-data-set analyses,
   including custom analysis functions, explicit delta data and parallel use.
5. Test conditional-mean bootstrap and jackknife layouts, including repeated
   source subjects, occurrence ID remapping, ordering and method counts. Test
   BMLMI blocks separately.
6. Test failures for inconsistent keys, duplicate occurrence/visit rows,
   factor level drift, remaining outcome missingness, invalid source IDs and
   incompatible method counts.
7. Update printing and document the unavailable-metadata behaviour.

The implementation should land with both fixed-grid Rubin imports and
conditional-mean resampling imports. BMLMI uses the same mapping machinery but
should only be enabled once its `B`-by-`D` ordering has dedicated tests.

## End-to-end examples

The following executable compatibility prototype uses rbmi's existing compact
imputation storage. Requiring `original` is a limitation of this short
prototype, not of the proposed completed-data backend. It deliberately uses
only existing rbmi method classes, so the imported object can be passed
directly to the unmodified `rbmi::analyse()`. Once `pooling_method()` is used
internally by `as_analysis()`, `method_unknown()` will follow exactly the same
path; it does not require a separate analysis function. After the package
implementation lands, this setup chunk can be removed and the examples can
call `rbmi::import()` directly.

```{r}
#| label: executable-import-prototype
#| warning: false

library(rbmi)

import <- function(
    data,
    vars,
    method,
    original,
    data_ice = NULL,
    references,
    source_id = NULL,
    ...
) {
    UseMethod("import", method)
}

validate_completed <- function(data, vars) {
    stopifnot(
        length(data) > 0L,
        all(vapply(data, is.data.frame, logical(1))),
        all(vapply(
            data,
            function(x) {
                all(c(vars$subjid, vars$visit, vars$group, vars$outcome) %in%
                    names(x)) &&
                    is.numeric(x[[vars$outcome]]) &&
                    !anyNA(x[[vars$outcome]]) &&
                    !anyDuplicated(x[c(vars$subjid, vars$visit)])
            },
            logical(1)
        ))
    )
}

validate_fixed_grid <- function(data, vars, expected) {
    validate_completed(data, vars)
    stopifnot(length(data) == expected)
    key <- function(x) {
        sort(paste(x[[vars$subjid]], x[[vars$visit]], sep = "\r"))
    }
    keys <- lapply(data, key)
    stopifnot(all(vapply(keys[-1L], identical, logical(1), keys[[1L]])))
}

import_compact <- function(
    data,
    vars,
    method,
    original,
    data_ice,
    references,
    source_id
) {
    longdata <- longDataConstructor$new(original, vars)
    longdata$set_strategies(data_ice)

    to_imputation_df <- function(x) {
        occurrences <- unique(as.character(x[[vars$subjid]]))
        singles <- lapply(occurrences, function(occurrence) {
            rows <- x[as.character(x[[vars$subjid]]) == occurrence, , drop = FALSE]
            source <- if (is.null(source_id)) {
                occurrence
            } else {
                unique(as.character(rows[[source_id]]))
            }
            stopifnot(length(source) == 1L, source %in% longdata$ids)

            visit_order <- match(
                longdata$visits,
                as.character(rows[[vars$visit]])
            )
            stopifnot(!anyNA(visit_order))
            values <- rows[[vars$outcome]][visit_order]
            rbmi:::imputation_single(
                id = source,
                values = values[longdata$is_missing[[source]]]
            )
        })
        rbmi:::imputation_df(singles)
    }

    stored <- lapply(data, to_imputation_df)
    stored <- do.call(rbmi:::imputation_list_df, stored)
    references <- add_class(references, "references")

    out <- rbmi:::as_imputation(
        imputations = stored,
        data = longdata,
        method = method,
        references = references
    )
    validate(out)
    out
}

import.bayes <- function(
    data, vars, method, original, data_ice = NULL, references,
    source_id = NULL, ...
) {
    validate_fixed_grid(data, vars, method$n_samples)
    import_compact(
        data, vars, method, original, data_ice, references, source_id
    )
}

import.approxbayes <- import.bayes

import.condmean <- function(
    data, vars, method, original, data_ice = NULL, references,
    source_id = NULL, ...
) {
    validate_completed(data, vars)
    stopifnot(
        method$type == "bootstrap",
        length(data) == method$n_samples + 1L,
        !is.null(source_id),
        all(vapply(data, function(x) source_id %in% names(x), logical(1)))
    )
    import_compact(
        data, vars, method, original, data_ice, references, source_id
    )
}
```

The first executed example imports ordinary Rubin-style MI data with a known
Bayesian pooling contract. In the implemented design, replacing
`method_bayes()` with `method_unknown("rubin")` changes only method metadata;
the returned imputation object and the call to `analyse()` are unchanged. The
second example shows the additional ID information required for
conditional-mean bootstrap data.

### Rubin-style MI

```{r}
#| label: example-rubin

set.seed(2026)
n_subjects <- 40L
n_imputations <- 5L

subject <- data.frame(
    USUBJID = factor(sprintf("P%03d", seq_len(n_subjects))),
    ARM = factor(rep(c("Placebo", "Active"), each = n_subjects / 2L)),
    BASE = rnorm(n_subjects, mean = 10, sd = 2)
)

original <- subject[rep(seq_len(n_subjects), each = 2L), ]
original$AVISIT <- factor(
    rep(c("Week 4", "Week 8"), times = n_subjects),
    levels = c("Week 4", "Week 8")
)
original$CHG <-
    -0.4 * original$BASE +
    1.5 * (original$ARM == "Active") +
    0.5 * (original$AVISIT == "Week 8") +
    rnorm(nrow(original))

missing_subjects <- sample(levels(original$USUBJID), 10L)
is_missing <-
    original$USUBJID %in% missing_subjects &
    original$AVISIT == "Week 8"
original$CHG[is_missing] <- NA_real_

# Stand-in for files received from an external imputation system.
completed <- lapply(seq_len(n_imputations), function(i) {
    out <- original
    out$CHG[is_missing] <-
        -0.4 * out$BASE[is_missing] +
        1.5 * (out$ARM[is_missing] == "Active") +
        0.5 +
        rnorm(sum(is_missing))
    out
})

vars <- set_vars(
    subjid = "USUBJID",
    visit = "AVISIT",
    group = "ARM",
    outcome = "CHG",
    covariates = "BASE"
)

references <- setNames(levels(original$ARM), levels(original$ARM))

imputations <- import(
    data = completed,
    vars = vars,
    method = method_bayes(n_samples = n_imputations),
    original = original,
    references = references
)

analysis <- rbmi::analyse(
    imputations = imputations,
    fun = ancova,
    vars = vars,
    visits = "Week 8"
)
pooled <- pool(analysis)
as.data.frame(pooled)
```

### Conditional-mean bootstrap

The original-sample conditional means are first in `completed_bootstrap`.
Every later element is one stratified bootstrap sample. `USUBJID` identifies a
unique sampled occurrence, while `SOURCE_ID` retains the original subject ID;
therefore a source subject selected twice is represented unambiguously.

```{r}
#| label: example-condmean-bootstrap

B <- 10L

# For illustration, create deterministic conditional-mean replacements.
conditional_mean <- original
conditional_mean$CHG[is_missing] <-
    -0.4 * conditional_mean$BASE[is_missing] +
    1.5 * (conditional_mean$ARM[is_missing] == "Active") +
    0.5
conditional_mean$SOURCE_ID <- as.character(conditional_mean$USUBJID)

arm_by_subject <- split(
    as.character(subject$USUBJID),
    subject$ARM
)

bootstrap_data <- lapply(seq_len(B), function(b) {
    sampled_sources <- unlist(lapply(
        arm_by_subject,
        function(ids) sample(ids, length(ids), replace = TRUE)
    ), use.names = FALSE)

    sampled_occurrences <- lapply(seq_along(sampled_sources), function(j) {
        source <- sampled_sources[[j]]
        out <- conditional_mean[
            as.character(conditional_mean$USUBJID) == source,
            ,
            drop = FALSE
        ]
        out$SOURCE_ID <- source
        out$USUBJID <- sprintf("B%02d_O%03d", b, j)
        out
    })

    out <- do.call(rbind, sampled_occurrences)
    rownames(out) <- NULL
    out$USUBJID <- factor(out$USUBJID)
    out
})

completed_bootstrap <- c(list(conditional_mean), bootstrap_data)

bootstrap_imputations <- import(
    data = completed_bootstrap,
    vars = vars,
    method = method_condmean(type = "bootstrap", n_samples = B),
    original = original,
    references = references,
    source_id = "SOURCE_ID"
)

bootstrap_analysis <- rbmi::analyse(
    imputations = bootstrap_imputations,
    fun = ancova,
    vars = vars,
    visits = "Week 8"
)
bootstrap_pooled <- pool(bootstrap_analysis)
as.data.frame(bootstrap_pooled)
```
