---
title: "rbmi: Modeling count data"
author: Daniel Sabanés Bové
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    number_sections: true
    citation_package: natbib
    base_format: rmarkdown::html_vignette
bibliography: "references.bib"
link-citations: true   
linkcolor: blue
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{rbmi: Modeling count data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

TODO: Write a nice introduction here, leveraging the design doc contents.

Now we'll walk through how count data can be modeled using the `rbmi` package. The following sections will cover the necessary steps to prepare your data, draw parameter samples, impute missing values, and pool estimates across multiple imputations.

## The Data

We are going to take the `epil` dataset from the `MASS` package as starting point and then wrangle it into the format we need for `rbmi`:

```{r, message = FALSE}
library(MASS)
library(dplyr)

dat_count <- epil |> 
    select(
        subject, 
        period, 
        count = y, 
        treatment = trt,
        baseline = base,
        age
    ) |> 
    mutate(
        subject = as.character(subject),
        duration = 2
    ) |> 
    filter(period %in% c(1, 2, 3))
```

Note that we do change the interpretation and meaning of the data now substantially. This is not supposed to be a realistic analysis of the `epil` data!

To make things more interesting, let's introduce missing values as well, according to these patterns:

- Pattern 0 (no missings): Stayed in the trial and received treatment. Then only period 1 is available, the other periods 2 and 3 have duration zero and missing counts.
- Pattern 1: Stopped treatment, but always stayed in the trial. Then we have duration zero for period 3.
- Pattern 2: Stopped first treatment, and then dropped out of the trial. Then all 3 periods are relevant, i.e. have positive duration, but period 3 has missing counts.
- Pattern 3: Stopped treatment and dropped out of the trial at the same time. Then period 2 has duration zero and missing counts, and period 3 has positive duration and also missing counts. (In general, period 3 always has missing counts.)

```{r}
set.seed(123)

subject_ids <- unique(dat_count$subject)

# For each patient, draw a pattern from 0 to 3.
pattern <- data.frame(
    subject = subject_ids,
    pattern = sample(
        0:3, 
        size = length(subject_ids), 
        replace = TRUE, 
        prob = c(0.5, 0.3, 0.1, 0.1)
    )
)

# Then change the data according to the pattern.
dat_count <- dat_count |> 
    left_join(pattern, by = "subject") |> 
    mutate(
        duration = case_when(
            pattern == 0 & period %in% c(2, 3) ~ 0,
            pattern == 1 & period == 3 ~ 0,
            pattern == 3 & period == 2 ~ 0,
            TRUE ~ duration
        ),
        count = case_when(
            pattern == 0 & period %in% c(2, 3) ~ NA,
            pattern == 1 & period == 3 ~ NA,
            pattern == 2 & period == 3 ~ NA,
            pattern == 3 & period %in% c(2, 3) ~ NA,
            TRUE ~ count
        )
    )
head(dat_count)
```

The `expand_locf()` function might come in handy when there are some periods not available yet in the data set. Say, we omit the rows with duration 0:

```{r}
dat_count_without_duration0 <- dat_count |> 
    filter(duration > 0)
```

Then we can use the function - after converting period to character:

```{r}
devtools::load_all()
dat_count_filled <- dat_count_without_duration0 |> 
    mutate(period = as.character(period)) |>
    expand_locf(
        subject = unique(dat_count_without_duration0$subject),
        period = as.character(1:3),
        vars = c("treatment", "baseline", "age", "pattern"),
        group = "subject",
        order = c("subject", "period")
    ) |> 
    mutate(
        duration = ifelse(is.na(duration), 0, duration)
    )
```

## Draws

The inputs to the `draws()` function are now:

- `data`: The data frame as prepared above.
- `data_ice`: The data with follow-up intercurrent events (ICE)imputation strategies for each patient
- `vars`: The variables to be used in the model
- `method`: Here we can only use `method_bayes()` for fully Bayesian multiple imputation for our count data
- `outcome` (new): The outcome type, which defaults to "continuous", but here we will choose "count"
- `data_treatment_ice` (new): The off-treatment imputation strategies for each patient

Note that for count data we do not need to provide the time points of the intercurrent events in `data_ice` (or `data_treatment_ice`), because these are calculated automatically from the `duration` variable in the data.

### Imputation strategies

Here we just use the same imputation strategy for all patients, but we would also be able to provide different strategies for each patient. 

Let's use the "Jump to Reference" (JR) strategy for all patients who discontinued the trial, i.e. those with pattern 2 or 3:

```{r}
dat_ice <- dat_count_filled |> 
    filter(pattern %in% c(2, 3)) |> 
    select(subject) |> 
    distinct() |> 
    mutate(
        strategy = "JR"
    )
```

For the off-treatment imputation strategy, we choose here the "On/Off Intercepts Common Slope" (OICS) strategy [@Rehal2026] for all patients who discontinued treatment, i.e. those with pattern 1 to 3:

```{r}
dat_treatment_ice <- dat_count_filled |> 
    filter(pattern %in% c(1, 2, 3)) |> 
    select(subject) |> 
    distinct() |> 
    mutate(
        strategy = "OICS"
    )
```

This means that in the imputation model we use treatment arm specific intercepts for the on- and off-treatment periods, but common coefficients for the other covariates. (We could also say we thus use an interaction term between treatment and the on/off-treatment indicator in the model.)

### Variables

Next we define the variables to be used in the model. 

```{r}
vars <- set_vars(
    outcome = "count",
    subjid = "subject",
    period = "period",
    duration = "duration",
    group = "treatment",
    covariates = c("baseline", "age")
)
```

## References 
