---
title: "rbmi: Modeling count data"
author: Daniel Sabanés Bové
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    number_sections: true
    citation_package: natbib
    base_format: rmarkdown::html_vignette
bibliography: "references.bib"
link-citations: true   
linkcolor: blue
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{rbmi: Modeling count data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

TODO: Write a nice introduction here, leveraging the design doc contents.

Now we'll walk through how count data can be modeled using the `rbmi` package. The following sections will cover the necessary steps to prepare your data, draw parameter samples, impute missing values, and pool estimates across multiple imputations.

## The Data

We are going to take the `epil` dataset from the `MASS` package as starting point and then wrangle it into the format we need for `rbmi`:

```{r, message = FALSE}
library(MASS)
library(dplyr)

dat_count <- epil |> 
    select(
        subject, 
        period, 
        count = y, 
        treatment = trt,
        baseline = base,
        age
    ) |> 
    mutate(
        subject = as.character(subject),
        duration = 2
    ) |> 
    filter(period %in% c(1, 2, 3))
```

Note that we do change the interpretation and meaning of the data now substantially. This is not supposed to be a realistic analysis of the `epil` data!

To make things more interesting, let's introduce missing values as well, according to these patterns:

- Pattern 0 (no missings): Stayed in the trial and received treatment. Then only period 1 is available, the other periods 2 and 3 have duration zero and missing counts.
- Pattern 1: Stopped treatment, but always stayed in the trial. Then we have duration zero for period 3.
- Pattern 2: Stopped first treatment, and then dropped out of the trial. Then all 3 periods are relevant, i.e. have positive duration, but period 3 has missing counts.
- Pattern 3: Stopped treatment and dropped out of the trial at the same time. Then period 2 has duration zero and missing counts, and period 3 has positive duration and also missing counts. (In general, period 3 always has missing counts.)

```{r}
set.seed(123)

subject_ids <- unique(dat_count$subject)

# For each patient, draw a pattern from 0 to 3.
pattern <- data.frame(
    subject = subject_ids,
    pattern = sample(
        0:3, 
        size = length(subject_ids), 
        replace = TRUE, 
        prob = c(0.5, 0.3, 0.1, 0.1)
    )
)

# Then change the data according to the pattern.
dat_count <- dat_count |> 
    left_join(pattern, by = "subject") |> 
    mutate(
        duration = case_when(
            pattern == 0 & period %in% c(2, 3) ~ 0,
            pattern == 1 & period == 3 ~ 0,
            pattern == 3 & period == 2 ~ 0,
            TRUE ~ duration
        ),
        count = case_when(
            pattern == 0 & period %in% c(2, 3) ~ NA,
            pattern == 1 & period == 3 ~ NA,
            pattern == 2 & period == 3 ~ NA,
            pattern == 3 & period %in% c(2, 3) ~ NA,
            TRUE ~ count
        )
    )
head(dat_count)
```

The `expand_locf()` function might come in handy when there are some periods not available yet in the data set. Say, we omit the rows with duration 0:

```{r}
dat_count_without_duration0 <- dat_count |> 
    filter(duration > 0)
```

Then we can use the function - after converting period to character:

```{r}
devtools::load_all()
dat_count_filled <- dat_count_without_duration0 |> 
    mutate(period = as.character(period)) |>
    expand_locf(
        subject = unique(dat_count_without_duration0$subject),
        period = as.character(1:3),
        vars = c("treatment", "baseline", "age", "pattern"),
        group = "subject",
        order = c("subject", "period")
    ) |> 
    mutate(
        duration = ifelse(is.na(duration), 0, duration)
    )
```