---
title: "rbmi: Modeling count data"
author: Daniel Sabanés Bové
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    number_sections: true
    citation_package: natbib
    base_format: rmarkdown::html_vignette
bibliography: "references.bib"
link-citations: true   
linkcolor: blue
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{rbmi: Modeling count data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

Now we'll walk through how count data can be modeled using the `rbmi` package. The following sections will cover the necessary steps to prepare your data, draw parameter samples, impute missing values, and pool estimates across multiple imputations.

## The Data

We use the simulated example data set from Roger et al. The data contain three treatment arms and up to three rows per patient, corresponding to the on-treatment, off-treatment but observed, and post-withdrawal periods.

```{r, message = FALSE}
library(dplyr)
pkgload::load_all()

dat_count <- read.csv(system.file("data", "sim_rogeretal.csv", package = "rbmi")) |>
    dplyr::select(
        patient,
        period,
        TreatLab,
        OnOff,
        Length,
        BaseCount,
        Observed_Count
    ) |> 
    mutate(
        patient = factor(patient),
        period = as.character(period),
        TreatLab = factor(TreatLab, levels = c("Placebo", "100mg", "300mg")),
        OnOff = factor(OnOff, levels = c("On", "Off"))
    )
head(dat_count)
```

The source data omit periods with zero length. We use `expand_locf()` to add these rows so that every patient has all three periods. The treatment arm and baseline count are carried forward, while `OnOff` is derived from the period. Added rows have zero length and a missing observed count.

```{r}
dat_count_filled <- dat_count |>
    expand_locf(
        patient = levels(dat_count$patient),
        period = as.character(1:3),
        vars = c("TreatLab", "BaseCount"),
        group = "patient",
        order = c("patient", "period")
    ) |> 
    mutate(
        OnOff = factor(
            ifelse(period == "1", "On", "Off"),
            levels = c("On", "Off")
        ),
        Length = ifelse(is.na(Length), 0, Length)
    )
```

## Draws

The inputs to the `draws()` function are now:

- `data`: The data frame as prepared above.
- `data_ice`: The data with follow-up intercurrent event (ICE) imputation strategies for each patient
- `vars`: The variables to be used in the model
- `method`: Here we can only use `method_bayes()` for fully Bayesian multiple imputation for our count data
- `outcome` (new): The outcome type, which defaults to "continuous", but here we will choose "count"

Note that for count data we do not need to provide the time points of the intercurrent events in `data_ice`, because these are calculated automatically from the `duration` variable in the data.

### Imputation strategies

Here we just use the same imputation strategy for all patients, but we would also be able to provide different strategies for each patient.

Let's use the Missing At Random (MAR) strategy for all patients with positive post-withdrawal exposure in period 3. This matches the first model in the "Another case with three periods" analysis, where the period-3 design matrix retains each patient's randomized treatment arm:

```{r}
dat_ice <- dat_count_filled |> 
    filter(period == "3", Length > 0) |>
    dplyr::select(patient) |>
    distinct() |> 
    mutate(
        strategy = "MAR"
    )
```

### Variables

Next we define the variables to be used in the model. 

```{r}
vars <- set_vars(
    outcome = "Observed_Count",
    subjid = "patient",
    period = "period",
    duration = "Length",
    group = "TreatLab",
    covariates = c("TreatLab * OnOff", "BaseCount")
)
```

Note that here we can also specify interaction terms in the `covariates` argument,
which will then be included in the draws step model.

### Method

We just use the standard Bayesian method options:

```{r}
method <- method_bayes(
    n_samples = 1000,
    control = control_bayes(
        warmup = 200,
        thin = 5,
        seed = 1821 # Seed to be used by Stan
    )
)
```

### Sampling

Finally we can call the `draws()` function to draw samples from the posterior distribution of the model parameters:

```{r}
set.seed(987)
drawObj <- draws(
    outcome = "count",
    data = dat_count_filled,
    data_ice = dat_ice,
    vars = vars,
    method = method,
    quiet = TRUE
)
drawObj
```

## Impute

Now we basically just run the imputation:

```{r}
imputeObj <- impute(drawObj)
imputeObj
```

## Analyse

Then the next step is the analysis of the imputed data. We can use a similar negative binomial regression model
as above, however here we don't have anymore the `OnOff` variable, because we have imputed the post-withdrawal
periods. So we just use baseline count as additional covariate, next to the treatment group.
The imputed values replace the missing values in the original outcome column. We therefore keep
`Observed_Count` as the outcome name. The analysis function sums this outcome and `Length` over periods for
each patient before fitting the model; it does not require a separate `Imputed_Count` column.

```{r}
ana_vars <- set_vars(
    outcome = "Observed_Count",
    subjid = "patient",
    period = "period",
    duration = "Length",
    group = "TreatLab",
    covariates = c("BaseCount")
)

anaObj <- analyse(
    imputeObj,
    neg_bin_regression,
    vars = ana_vars
)
```

## Pool

Finally we can pool all regression coefficients across the multiple
imputations. Pooling is performed on the log-rate scale using Rubin's rules,
with confidence intervals and p-values based on the standard normal
distribution.

```{r}
poolObj <- pool(
    anaObj,
    conf.level = 0.95,
    alternative = "two.sided"
)
poolObj
```

Treatment-group coefficients and their confidence limits can be exponentiated
to obtain rate ratios. Other coefficients should only be exponentiated when
that interpretation is appropriate for the corresponding model term.

## Additional modeling options

### Separate draw model sampling for each treatment group

The default count model uses one dispersion/frailty parameter across all
treatment groups. To estimate a separate parameter for each group, set
`same_cov = FALSE` in `method_bayes()`. The name is retained for consistency
with continuous-outcome models; for count outcomes it controls whether `phi`
is shared between groups.

To make the complete imputation model separate by treatment arm, all fixed
effects must also be interacted with treatment. For the example data this is:

```{r}
vars_by_arm <- set_vars(
    outcome = "Observed_Count",
    subjid = "patient",
    period = "period",
    duration = "Length",
    group = "TreatLab",
    covariates = c(
        "TreatLab:OnOff",
        "TreatLab:BaseCount"
    )
)

method_by_arm <- method_bayes(
    same_cov = FALSE,
    n_samples = 1000,
    control = control_bayes(
        warmup = 200,
        thin = 5,
        seed = 1821
    )
)

drawObj_by_arm <- draws(
    outcome = "count",
    data = dat_count_filled,
    data_ice = dat_ice,
    vars = vars_by_arm,
    method = method_by_arm,
    quiet = TRUE
)
```

This is fitted in one Stan call, but with treatment-specific fixed effects and
independent treatment-specific `phi` parameters the posterior factorizes by
arm. It therefore represents the same statistical model as fitting each arm
separately, while retaining a single rbmi draws object and the usual downstream
workflow.

For MAR imputation, each subject uses the `phi` for their randomized treatment
arm. For JR and CR, the mapped reference arm's `phi` is used, consistently with
the reference-arm model used for imputation.

#### Impute

The downstream workflow is unchanged. Setting the seed immediately before
`impute()` makes the conditional negative-binomial draws reproducible.

```{r}
set.seed(123)
imputeObj_by_arm <- impute(drawObj_by_arm)
```

#### Analyse and pool

The treatment-policy analysis still combines all randomized treatment arms and
uses the same analysis model as above. Only the imputation model has changed.

```{r}
anaObj_by_arm <- analyse(
    imputeObj_by_arm,
    neg_bin_regression,
    vars = ana_vars
)

poolObj_by_arm <- pool(
    anaObj_by_arm,
    conf.level = 0.95,
    alternative = "two.sided"
)
poolObj_by_arm
```

For comparison with implementations that report treatment effects as rate
ratios, exponentiate the treatment coefficients and their confidence limits:

```{r}
rate_ratios_by_arm <- as.data.frame(poolObj_by_arm) |>
    filter(parameter %in% c("TreatLab100mg", "TreatLab300mg")) |>
    transmute(
        comparison = recode(
            parameter,
            TreatLab100mg = "100mg vs Placebo",
            TreatLab300mg = "300mg vs Placebo"
        ),
        rate_ratio = exp(est),
        lower_95 = exp(lci),
        upper_95 = exp(uci),
        p_value = pval
    )
rate_ratios_by_arm
```

## References 
